# Fennica: Nineteenth-Century Publication Metadata (1809–1917)

This chapter analyzes historical records retrieved from the Finnish National
Bibliography (Fennica) through the Finna ecosystem. It characterizes
publication growth over time, language distributions and language change,
publication formats, and metadata completeness and quality.

> **Data path note.** This chapter reads `nineteen_cen.rds` from
> `~/finna/inst/extras/`. Update the path in the chunk below if your local
> copy lives elsewhere.

## Setup

```{r}
#| label: setup-fennica

library(dplyr)
library(ggplot2)
library(scales)
library(forcats)
library(patchwork)
library(stringr)
library(tidyr)
library(grid)
```

## Load harmonized Fennica metadata

The dataset contains historical Fennica records covering the
nineteenth-century publication landscape in Finland.

```{r}
#| label: load-data-fennica

nineteen_cen <- readRDS("~/finna/inst/extras/nineteen_cen.rds")
```

## Translation dictionaries

Finnish language names and ISO language codes are converted into
standardized English labels to improve figure readability.

```{r}
#| label: translation-maps-fennica

# Human-readable language names
lang_map <- c(
  "suomi"    = "Finnish",
  "ruotsi"   = "Swedish",
  "englanti" = "English",
  "saksa"    = "German",
  "ranska"   = "French",
  "venäjä"   = "Russian",
  "latina"   = "Latin"
)

# ISO 639 language codes
lang_code_map <- c(
  "fin" = "Finnish",
  "swe" = "Swedish",
  "rus" = "Russian",
  "deu" = "German",
  "ger" = "German",
  "eng" = "English",
  "fre" = "French",
  "fra" = "French",
  "lat" = "Latin",
  "est" = "Estonian"
)

# Combined lookup table
language_map <- c(lang_map, lang_code_map)

# Finnish format labels -> English labels
format_map <- c(
  "kirja"           = "Book",
  "lehti/artikkeli" = "Journal article",
  "opinnäyte"       = "Thesis",
  "kartta"          = "Map",
  "muu"             = "Other",
  "peli"            = "Game",
  "nuotti"          = "Sheet music",
  "teksti"          = "Text",
  "kuva"            = "Image"
)
```

## Helper function

Standardizes metadata labels by trimming whitespace, lower-casing, and
translating known labels while preserving unknown ones.

```{r}
#| label: helper-fennica

label_en <- function(x, map) {
  x0 <- tolower(trimws(x))
  out <- unname(map[x0])
  ifelse(is.na(out), x0, out)
}
```

## Plot styling

A muted colour palette and shared theme are used across all figures to
ensure consistent, publication-ready styling.

```{r}
#| label: styling-fennica

quiet_pal <- c(
  "Finnish" = "#2B2B2B",
  "Swedish" = "#D89000",
  "Other"   = "#9E9E9E",
  "Russian" = "#6BAA75",
  "German"  = "#D8C94A"
)

base_theme <- theme_minimal(base_size = 14) +
  theme(
    plot.title   = element_text(face = "bold", size = 15, margin = margin(b = 8)),
    axis.title   = element_text(size = 14),
    axis.text    = element_text(size = 13),
    legend.title = element_blank(),
    legend.text  = element_text(size = 14),
    plot.margin  = margin(t = 12, r = 14, b = 12, l = 28)
  )
```

## Data preparation

Records lacking language or year information are removed, publication year
is converted to numeric, decade is derived, language values are
harmonized, and rare language categories are merged into "Other".

```{r}
#| label: data-prep-fennica

nineteen_cen_clean <- nineteen_cen %>%
  filter(!is.na(Language), !is.na(Year)) %>%
  mutate(
    Year = suppressWarnings(as.numeric(Year)),
    Decade = floor(Year / 10) * 10,
    Language_raw = tolower(trimws(Language)),
    Language_lab = label_en(Language_raw, language_map),
    Language_lab = ifelse(
      tolower(Language_lab) %in% c("other", "others", "muu", "unknown", ""),
      "Other",
      Language_lab
    )
  ) %>%
  filter(!is.na(Year))

# Five most common publication languages
top_langs <- nineteen_cen_clean %>%
  count(Language_lab, sort = TRUE) %>%
  slice_head(n = 5) %>%
  pull(Language_lab)

# Collapse remaining languages into "Other"
nineteen_cen_clean <- nineteen_cen_clean %>%
  mutate(
    Language = ifelse(Language_lab %in% top_langs, Language_lab, "Other")
  )
```

## Publication growth and language trends

### Publications by decade {#sec-fennica-fig1a}

Logarithmic scaling is used because publication counts increase
substantially through the nineteenth century.

```{r}
#| label: fig-fennica-decade
#| fig-cap: "Publications in Fennica by decade (log scale)"

p1 <- nineteen_cen_clean %>%
  group_by(Decade) %>%
  summarise(Publications = n(), .groups = "drop") %>%
  ggplot(aes(x = Decade, y = Publications)) +
  geom_col(fill = "grey35") +
  scale_y_log10(labels = comma) +
  labs(
    title = "Publications in Fennica by decade (log scale)",
    x = "Decade",
    y = "Publications"
  ) +
  base_theme

p1
```

### Annual publication counts {#sec-fennica-fig1b}

A LOESS smoother emphasizes long-term trends while reducing short-term
fluctuations.

```{r}
#| label: fig-fennica-annual
#| fig-cap: "Annual publications in Fennica (log scale)"

p2 <- nineteen_cen_clean %>%
  group_by(Year) %>%
  summarise(Publications = n(), .groups = "drop") %>%
  ggplot(aes(x = Year, y = Publications)) +
  geom_line(color = "grey35", linewidth = 0.5) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    linetype = "dashed",
    color = "grey55"
  ) +
  scale_y_log10(labels = comma) +
  labs(
    title = "Annual publications in Fennica (log scale)",
    x = "Year",
    y = "Publications"
  ) +
  base_theme

p2
```

### Most common publication languages {#sec-fennica-fig1c}

```{r}
#| label: fig-fennica-top-languages
#| fig-cap: "Most common publication languages"

p3 <- nineteen_cen_clean %>%
  count(Language, sort = TRUE) %>%
  ggplot(aes(x = fct_reorder(Language, n), y = n)) +
  geom_col(fill = "grey35") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Most common publication languages",
    x = "Language",
    y = "Number of publications"
  ) +
  base_theme

p3
```

### Language trend calculations

Publication counts are aggregated by decade and language for reuse in the
composition and proportion figures below.

```{r}
#| label: lang-trends-fennica

lang_trends <- nineteen_cen_clean %>%
  group_by(Decade, Language) %>%
  summarise(Count = n(), .groups = "drop")

lang_levels <- lang_trends %>%
  group_by(Language) %>%
  summarise(Total = sum(Count), .groups = "drop") %>%
  arrange(desc(Total)) %>%
  pull(Language)

lang_trends <- lang_trends %>%
  mutate(Language = factor(Language, levels = lang_levels))
```

### Language composition through time {#sec-fennica-fig1d}

Absolute publication counts by language for each decade, useful for
examining growth and decline of individual language groups.

```{r}
#| label: fig-fennica-language-composition
#| fig-cap: "Language composition by decade"

p4 <- ggplot(lang_trends, aes(x = Decade, y = Count, fill = Language)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(
    values = rep(quiet_pal, length.out = nlevels(lang_trends$Language))
  ) +
  labs(
    title = "Language composition by decade",
    x = "Decade",
    y = "Number of publications"
  ) +
  base_theme +
  theme(legend.position = "none")

p4
```

### Relative language proportions {#sec-fennica-fig1e}

Changing language shares independent of collection size, highlighting
shifts in linguistic dominance over time.

```{r}
#| label: fig-fennica-language-proportions
#| fig-cap: "Language proportions by decade"

lang_share <- lang_trends %>%
  group_by(Decade) %>%
  mutate(Proportion = Count / sum(Count)) %>%
  ungroup()

p5 <- ggplot(lang_share, aes(x = Decade, y = Proportion, fill = Language)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(
    values = rep(quiet_pal, length.out = nlevels(lang_trends$Language))
  ) +
  labs(
    title = "Language proportions by decade",
    x = "Decade",
    y = "Proportion of publications"
  ) +
  base_theme +
  theme(legend.position = "bottom")

p5
```

### Most common publication formats {#sec-fennica-fig1f}

Log scale is used because format frequencies differ substantially.

```{r}
#| label: fig-fennica-top-formats
#| fig-cap: "Top five formats in Fennica (log scale)"

p6 <- nineteen_cen_clean %>%
  filter(!is.na(Formats)) %>%
  mutate(
    Format_raw = tolower(trimws(gsub(",.*", "", Formats))),
    Format = label_en(Format_raw, format_map),
    Format = ifelse(
      tolower(Format) %in% c("other", "others", "muu", "unknown", ""),
      "Other",
      Format
    )
  ) %>%
  count(Format, sort = TRUE) %>%
  slice_head(n = 5) %>%
  ggplot(aes(x = fct_reorder(Format, n), y = n)) +
  geom_col(fill = "grey35") +
  coord_flip() +
  scale_y_log10(labels = comma) +
  labs(
    title = "Top five formats in Fennica (log scale)",
    x = "Format",
    y = "Publications"
  ) +
  base_theme

p6
```

### Combined six-panel figure

A publication-ready multi-panel figure combining publication growth,
language, and format trends.

```{r}
#| label: fig-fennica-combined-6panel
#| fig-cap: "Fennica metadata analysis, 1809–1917"
#| fig-width: 16
#| fig-height: 12

combined_plot_6 <-
  (p1 + p2) /
  (p3 + p4) /
  (p5 + p6) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Fennica metadata analysis, 1809–1917",
    tag_levels = "a",
    theme = theme(
      plot.title = element_text(
        size = 20,
        face = "bold",
        margin = margin(b = 18)
      )
    )
  ) &
  theme(
    legend.position = "bottom",
    plot.tag = element_text(face = "bold", size = 16),
    plot.title = element_text(face = "bold")
  )

combined_plot_6
```

```{r}
#| label: save-fennica-6panel
#| eval: false

ggsave(
  "fennica_combined_6panel_en.png",
  combined_plot_6,
  width = 16,
  height = 12,
  units = "in",
  dpi = 300
)
```

## Metadata quality analysis

This section examines metadata coverage, missingness, and the interaction
between publication format and language.

### Additional cleaning

Publication format is extracted and key metadata fields (Author, Language,
Subjects, Library, Series) are standardized.

```{r}
#| label: data-prep-quality-fennica

fennica_clean <- nineteen_cen %>%
  mutate(
    Year     = suppressWarnings(as.integer(Year)),
    Decade   = (Year %/% 10) * 10,
    Language = tolower(trimws(Language)),
    Formats  = trimws(Formats),
    Format   = trimws(sub("\\s*[,;|].*$", "", Formats))
  ) %>%
  filter(!is.na(Year), Year >= 1809, Year <= 1917)

fc <- fennica_clean %>%
  mutate(
    Year     = suppressWarnings(as.numeric(Year)),
    Decade   = floor(Year / 10) * 10,
    Language = na_if(trimws(Language), ""),
    Format   = na_if(trimws(Format), ""),
    Author   = na_if(trimws(Author), ""),
    Subjects = na_if(trimws(Subjects), ""),
    Library  = na_if(trimws(Library), ""),
    Series   = na_if(trimws(Series), "")
  ) %>%
  filter(!is.na(Year), !is.na(Decade))

format_map <- c(
  "kirja" = "Book",
  "lehti/artikkeli" = "Journal article",
  "opinnäyte" = "Thesis",
  "kartta" = "Map",
  "muu" = "Other",
  "peli" = "Game",
  "nuotti" = "Sheet music",
  "teksti" = "Text",
  "kuva" = "Image"
)

lang_map <- c(
  "fin" = "Finnish",
  "swe" = "Swedish",
  "eng" = "English",
  "deu" = "German",
  "ger" = "German",
  "rus" = "Russian",
  "fra" = "French",
  "fre" = "French",
  "lat" = "Latin",
  "zxx" = "No linguistic content"
)

label_en <- function(x, map) {
  x0 <- tolower(trimws(x))
  out <- unname(map[x0])
  ifelse(is.na(out), x, out)
}

fc <- fc %>%
  mutate(
    Format   = ifelse(is.na(Format), NA_character_, label_en(Format, format_map)),
    Language = ifelse(is.na(Language), NA_character_, label_en(Language, lang_map))
  )
```

### Colour schemes and theme

Neutral colours are used here to emphasize metadata quality patterns
rather than categorical distinctions.

```{r}
#| label: styling-quality-fennica

field_cols <- c(
  "Author name" = "#404040",
  "Language"    = "#7A7A7A",
  "Series"      = "#A6A6A6",
  "Subjects"    = "#D0D0D0"
)

heatmap_cols <- c(
  "#F7F7F7",
  "#D9D9D9",
  "#969696",
  "#525252",
  "#252525"
)

base_theme <- theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 11),
    axis.text = element_text(size = 10),
    legend.title = element_blank(),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.25, colour = "grey88"),
    panel.grid.major.y = element_line(linewidth = 0.25, colour = "grey88"),
    plot.margin = margin(8, 8, 8, 8)
  )
```

### Metadata coverage over time {#sec-fennica-fig2a}

Coverage is calculated as the share of records with each field present,
per decade.

```{r}
#| label: fig-fennica-coverage
#| fig-cap: "Metadata field coverage over time"

field_cov <- fc %>%
  group_by(Decade) %>%
  summarise(
    `Author name` = mean(!is.na(Author)),
    Language      = mean(!is.na(Language)),
    Series        = mean(!is.na(Series)),
    Subjects      = mean(!is.na(Subjects)),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(`Author name`, Language, Series, Subjects),
    names_to = "Field",
    values_to = "SharePresent"
  )

pA <- ggplot(field_cov, aes(Decade, SharePresent, colour = Field)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = field_cols) +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.25)
  ) +
  scale_x_continuous(breaks = seq(1810, 1910, 20)) +
  labs(
    title = "A. Metadata field coverage over time",
    subtitle = "Share of records with each field present",
    x = "Decade",
    y = "Share present"
  ) +
  base_theme

pA
```

### Metadata missingness {#sec-fennica-fig2b}

Directly quantifies missing values, complementing the coverage analysis
above.

```{r}
#| label: fig-fennica-missingness
#| fig-cap: "Field-level missingness by decade"

miss_long <- fc %>%
  group_by(Decade) %>%
  summarise(
    `Author name` = mean(is.na(Author)),
    Language      = mean(is.na(Language)),
    Series        = mean(is.na(Series)),
    Subjects      = mean(is.na(Subjects)),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(`Author name`, Language, Series, Subjects),
    names_to = "Field",
    values_to = "MissingRate"
  ) %>%
  mutate(Decade_f = factor(Decade))

pB <- ggplot(miss_long, aes(Decade_f, MissingRate, fill = Field)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.72
  ) +
  scale_fill_manual(values = field_cols) +
  scale_y_continuous(
    labels = percent_format(),
    limits = c(0, 1),
    breaks = seq(0, 1, 0.25)
  ) +
  labs(
    title = "B. Field-level missingness",
    subtitle = "Missing rate by decade",
    x = "Decade",
    y = "Missing rate"
  ) +
  base_theme +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )

pB
```

### Language–format interaction heatmap {#sec-fennica-fig2c}

Each cell represents the proportion of a language within a format
category.

```{r}
#| label: fig-fennica-heatmap
#| fig-cap: "Language distribution by format"

lf <- fc %>%
  filter(!is.na(Language), !is.na(Format)) %>%
  mutate(
    Format = fct_lump_n(factor(Format), n = 8, other_level = "Other"),
    Language = fct_lump_n(factor(Language), n = 5, other_level = "Other")
  )

prop_df <- as.data.frame(prop.table(table(lf$Language, lf$Format), margin = 2))
colnames(prop_df) <- c("Language", "Format", "share")

pC <- ggplot(prop_df, aes(Format, Language, fill = share)) +
  geom_tile(colour = "white", linewidth = 0.25) +
  scale_fill_gradientn(
    colours = heatmap_cols,
    labels = percent_format(),
    breaks = c(0, 0.25, 0.50, 0.75, 1),
    limits = c(0, 1),
    guide = guide_colorbar(
      title = "Share within format",
      title.position = "top",
      barwidth = unit(7, "cm"),
      barheight = unit(0.35, "cm")
    )
  ) +
  labs(
    title = "C. Language distribution by format",
    subtitle = "Cell value = share within each format",
    x = "Format",
    y = "Language"
  ) +
  base_theme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 10),
    legend.position = "bottom"
  )

pC
```

### Combined metadata quality figure

Combines coverage over time (A), missingness by decade (B), and language
distribution across formats (C) into a single overview figure.

```{r}
#| label: fig-fennica-quality-combined
#| fig-cap: "Fennica (1809–1917): Format interaction and metadata quality"
#| fig-width: 14
#| fig-height: 9

fig3 <- (pA) / (pB + pC) +
  plot_layout(heights = c(1.05, 1)) +
  plot_annotation(
    title = "Fennica (1809–1917): Format interaction and metadata quality"
  ) &
  theme(
    plot.title = element_text(face = "bold", size = 16)
  )

fig3
```

```{r}
#| label: save-fennica-quality
#| eval: false

ggsave("fennica_fig3_revised_minimal.png", fig3, width = 14, height = 9, dpi = 300)
```
