# Viola: Music Collection Metadata (1900–2025)

This chapter explores metadata from the Viola collection, the Finnish
national discography and bibliography of sheet music. It examines growth
in record counts by year and decade, publication language distributions
and language change, and compares major music-related formats — sound
recordings, sheet music, and video.

> **Data path note.** This chapter reads `viola_collection.rds` from
> `~/finna/inst/extras/`. Update the path in the chunk below if your local
> copy lives elsewhere.

## Setup

```{r}
#| label: setup-viola

library(dplyr)
library(ggplot2)
library(scales)
library(patchwork)
library(tidyr)
library(forcats)
```

## Load Viola metadata

```{r}
#| label: load-data-viola

viola_collection <- readRDS("~/finna/inst/extras/viola_collection.rds")
```

## Translation dictionaries

Language codes and Finnish language names are converted into English
labels, and Finnish format labels into English labels, for readability in
an English-language manuscript.

```{r}
#| label: translation-maps-viola

lang_code_map <- c(
  "fin" = "Finnish",
  "swe" = "Swedish",
  "eng" = "English",
  "deu" = "German",
  "ger" = "German",
  "rus" = "Russian",
  "fra" = "French",
  "fre" = "French",
  "lat" = "Latin",
  "est" = "Estonian",
  "zxx" = "No linguistic \n content"
)

lang_fi_map <- c(
  "suomi"    = "Finnish",
  "ruotsi"   = "Swedish",
  "englanti" = "English",
  "saksa"    = "German",
  "ranska"   = "French",
  "venäjä"   = "Russian",
  "latina"   = "Latin"
)

language_map <- c(lang_code_map, lang_fi_map)

# Only the most relevant Viola formats are included here
format_map <- c(
  "äänite"          = "Sound recording",
  "nuotti"          = "Sheet music",
  "video"           = "Video",
  "muu"             = "Other",
  "kirja"           = "Book",
  "teksti"          = "Text",
  "opinnäyte"       = "Thesis",
  "lehti/artikkeli" = "Journal article",
  "kuva"            = "Image"
)
```

## Helper function

Unknown values are kept in cleaned lowercase form.

```{r}
#| label: helper-viola

label_en <- function(x, map) {
  x0 <- tolower(trimws(x))
  out <- unname(map[x0])
  ifelse(is.na(out), x0, out)
}
```

## Styling

```{r}
#| label: styling-viola

lang_pal <- c(
  "Finnish" = "#000000",
  "Swedish" = "#E69F00",
  "English" = "#0072B2",
  "German"  = "#009E73",
  "Russian" = "#56B4E9",
  "No linguistic content" = "#CC79A7",
  "Other"   = "#7F7F7F"
)

format_pal <- c(
  "Sound recording" = "grey35",
  "Sheet music"     = "grey70",
  "Video"           = "#0072B2",
  "Other"           = "#7F7F7F"
)

# Short count labels prevent scientific notation
count_labels <- label_number(scale_cut = cut_short_scale())

base_theme <- theme_minimal(base_size = 17) +
  theme(
    plot.title   = element_text(face = "bold", size = 18),
    axis.title   = element_text(size = 16),
    axis.text    = element_text(size = 15),
    legend.title = element_blank(),
    legend.text  = element_text(size = 15),
    plot.margin  = margin(8, 8, 8, 8)
  )
```

## Data preparation

The analysis is restricted to records from 1900 onwards, with 2025 used
as the upper bound. Year and format labels are harmonized, and unknown or
rare values are standardized as "Other".

```{r}
#| label: data-prep-viola

start_year <- 1900
end_year <- 2025

viola_clean <- viola_collection %>%
  mutate(
    Year = suppressWarnings(as.numeric(Year)),
    Decade = floor(Year / 10) * 10,

    Language_raw = tolower(trimws(Language)),
    Language_lab = label_en(Language_raw, language_map),
    Language_lab = ifelse(
      tolower(Language_lab) %in% c("other", "others", "muu", "unknown", ""),
      "Other",
      Language_lab
    ),
    Format_raw = tolower(trimws(gsub(",.*", "", Formats))),
    Format_lab = label_en(Format_raw, format_map),
    Format_lab = ifelse(
      tolower(Format_lab) %in% c("other", "others", "muu", "unknown", ""),
      "Other",
      Format_lab
    )
  ) %>%
  filter(!is.na(Year), Year >= start_year, Year <= end_year)

# Five most frequent languages after 1900; remainder grouped as "Other"
top_langs <- viola_clean %>%
  filter(!is.na(Language_lab), Language_lab != "") %>%
  count(Language_lab, sort = TRUE) %>%
  slice_head(n = 5) %>%
  pull(Language_lab)

viola_clean <- viola_clean %>%
  mutate(
    Language = ifelse(Language_lab %in% top_langs, Language_lab, "Other"),
    Language = factor(Language, levels = c(top_langs, "Other"))
  )

# Maximum decade, to keep the x-axis range consistent
decade_max <- ceiling(max(viola_clean$Year, na.rm = TRUE) / 10) * 10
```

## Collection trends and language analysis

### Entries by decade {#sec-viola-panel1}

```{r}
#| label: fig-viola-decade
#| fig-cap: "Entries in Viola by decade, 1900–2025"

p1 <- viola_clean %>%
  count(Decade) %>%
  ggplot(aes(x = Decade, y = n)) +
  geom_col(fill = "grey40") +
  scale_y_continuous(labels = count_labels) +
  labs(
    title = "Entries in Viola by decade, 1900–2025",
    x = "Decade",
    y = "Count"
  ) +
  base_theme

p1
```

### Annual entries {#sec-viola-panel2}

The dashed LOESS curve highlights the longer-term trend.

```{r}
#| label: fig-viola-annual
#| fig-cap: "Annual entries in Viola, 1900–2025"

p2 <- viola_clean %>%
  count(Year) %>%
  ggplot(aes(x = Year, y = n)) +
  geom_line(color = "grey35", linewidth = 0.8) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    linetype = "dashed",
    color = "grey55",
    linewidth = 0.9
  ) +
  scale_y_continuous(labels = count_labels) +
  labs(
    title = "Annual entries in Viola, 1900–2025",
    x = "Year",
    y = "Count"
  ) +
  base_theme

p2
```

### Top publication languages {#sec-viola-panel3}

```{r}
#| label: fig-viola-top-languages
#| fig-cap: "Top publication languages in Viola"

p3 <- viola_clean %>%
  count(Language, sort = TRUE) %>%
  ggplot(aes(x = fct_reorder(Language, n), y = n)) +
  geom_col(fill = "grey40") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Top publication languages",
    x = "Language",
    y = "Count"
  ) +
  base_theme

p3
```

### Language shifts over time {#sec-viola-panel4}

Absolute language counts by decade, highlighting how dominant languages
change over time.

```{r}
#| label: fig-viola-language-shifts
#| fig-cap: "Language shifts over time in Viola"

p4 <- viola_clean %>%
  count(Decade, Language) %>%
  ggplot(aes(x = Decade, y = n, fill = Language)) +
  geom_col() +
  scale_y_continuous(labels = count_labels) +
  scale_fill_manual(values = lang_pal[levels(viola_clean$Language)]) +
  labs(
    title = "Language shifts over time",
    x = "Decade",
    y = "Count"
  ) +
  base_theme +
  theme(legend.position = "none")

p4
```

### Language proportions by decade {#sec-viola-panel5}

Relative language composition within each decade, removing the effect of
changing total record counts over time.

```{r}
#| label: fig-viola-language-proportions
#| fig-cap: "Language proportions by decade in Viola"

p5 <- viola_clean %>%
  count(Decade, Language) %>%
  group_by(Decade) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = Decade, y = Proportion, fill = Language)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = lang_pal[levels(viola_clean$Language)]) +
  scale_x_continuous(limits = c(min(viola_clean$Decade), decade_max)) +
  labs(title = "Language proportions by decade", x = "Decade", y = "Proportion") +
  base_theme +
  theme(legend.position = "bottom")

p5
```

### Format composition over time {#sec-viola-panel6}

The balance of major Viola formats by decade — more informative than a
simple total format-count plot, since it reveals structural change in the
collection over time.

```{r}
#| label: fig-viola-format-composition
#| fig-cap: "Format composition by decade in Viola"

top_formats <- viola_clean %>%
  filter(!is.na(Format_lab), Format_lab != "") %>%
  count(Format_lab, sort = TRUE) %>%
  slice_head(n = 3) %>%
  pull(Format_lab)

p6 <- viola_clean %>%
  filter(!is.na(Format_lab), Format_lab != "") %>%
  mutate(
    Format_group = ifelse(Format_lab %in% top_formats, Format_lab, "Other"),
    Format_group = factor(Format_group, levels = c(top_formats, "Other"))
  ) %>%
  count(Decade, Format_group) %>%
  group_by(Decade) %>%
  mutate(Proportion = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = Decade, y = Proportion, fill = Format_group)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_manual(values = format_pal[levels(
    factor(c(top_formats, "Other"), levels = c(top_formats, "Other"))
  )]) +
  labs(
    title = "Format composition by decade",
    x = "Decade",
    y = "Proportion"
  ) +
  base_theme +
  theme(legend.position = "bottom")

p6
```

### Combined six-panel figure

```{r}
#| label: fig-viola-combined-6panel
#| fig-cap: "Viola collection trends and language analysis, 1900–2025"
#| fig-width: 15
#| fig-height: 12

viola_combined <-
  (p1 + p2) /
  (p3 + p4) /
  (p5 + p6) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Viola collection trends and language analysis, 1900–2025"
  ) &
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 23, face = "bold")
  )

viola_combined
```

```{r}
#| label: save-viola-6panel
#| eval: false

ggsave(
  "viola_6panel_1900_onwards.png",
  viola_combined,
  width = 15,
  height = 12,
  dpi = 300
)
```

## Format-specific music metadata analysis after 1900

This section focuses specifically on the three core music-related
formats — sheet music, sound recordings, and video — comparing language
dynamics within each major musical material type.

### Translation dictionaries (repeated)

Repeated here so this section can be run independently if needed. Only
the three target formats are retained.

```{r}
#| label: translation-maps-viola-music

lang_code_map <- c(
  "fin" = "Finnish",
  "swe" = "Swedish",
  "eng" = "English",
  "deu" = "German",
  "ger" = "German",
  "rus" = "Russian",
  "fra" = "French",
  "fre" = "French",
  "lat" = "Latin",
  "est" = "Estonian",
  "zxx" = "No linguistic \n content"
)

lang_fi_map <- c(
  "suomi"    = "Finnish",
  "ruotsi"   = "Swedish",
  "englanti" = "English",
  "saksa"    = "German",
  "ranska"   = "French",
  "venäjä"   = "Russian",
  "latina"   = "Latin"
)

language_map <- c(lang_code_map, lang_fi_map)

format_map <- c(
  "äänite" = "Sound recording",
  "nuotti" = "Sheet music",
  "video"  = "Video"
)

label_en <- function(x, map) {
  x0 <- tolower(trimws(x))
  out <- unname(map[x0])
  ifelse(is.na(out) | out == "", x0, out)
}

quiet_pal <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2")

base_theme <- theme_minimal(base_size = 16) +
  theme(
    plot.title   = element_text(face = "bold", size = 16),
    axis.title   = element_text(size = 10),
    axis.text    = element_text(size = 10),
    legend.title = element_blank(),
    legend.text  = element_text(size = 16),
    strip.text   = element_text(size = 16, face = "bold"),
    plot.margin  = margin(8, 8, 8, 8)
  )
```

### Data preparation for format-specific analysis

Records are restricted to those with year and format information, after
1900, and limited to sheet music, sound recordings, and video.

```{r}
#| label: data-prep-viola-music

viola <- viola_collection %>%
  filter(!is.na(Year), !is.na(Formats)) %>%
  mutate(
    Year   = suppressWarnings(as.numeric(Year)),
    Decade = floor(Year / 10) * 10,
    Format_raw   = tolower(trimws(gsub(",.*", "", Formats))),
    Format       = label_en(Format_raw, format_map),
    Language_raw = tolower(trimws(Language)),
    Language_lab = label_en(Language_raw, language_map)
  ) %>%
  filter(!is.na(Year), Year >= 1900) %>%
  filter(Format %in% c("Sheet music", "Sound recording", "Video")) %>%
  mutate(
    Format = factor(Format, levels = c("Sheet music", "Sound recording", "Video"))
  )

# Six most common languages for the composition plot; remainder -> "Other"
top_langs_b <- viola %>%
  count(Language_lab, sort = TRUE) %>%
  slice_head(n = 6) %>%
  pull(Language_lab)

viola_b <- viola %>%
  mutate(Language = ifelse(Language_lab %in% top_langs_b, Language_lab, "Other"))

# Top 3 languages overall (post-1900) for panel (c)
top3_langs <- viola %>%
  count(Language_lab, sort = TRUE) %>%
  slice_head(n = 3) %>%
  pull(Language_lab)
```

### Format expansion after 1900 {#sec-viola-music-panel-a}

Annual growth patterns of sheet music, sound recordings, and video
records. Missing year-format combinations are filled with zero so each
format has a continuous time series.

```{r}
#| label: fig-viola-format-expansion
#| fig-cap: "Format expansion after 1900"

year_grid <- seq(
  min(viola$Year, na.rm = TRUE),
  max(viola$Year, na.rm = TRUE),
  by = 1
)

p1 <- viola %>%
  count(Year, Format, name = "n") %>%
  complete(Format, Year = year_grid, fill = list(n = 0)) %>%
  ggplot(aes(Year, n)) +
  geom_line(color = "grey35", linewidth = 0.7) +
  geom_smooth(
    method = "loess",
    se = FALSE,
    linetype = "dashed",
    color = "grey55",
    linewidth = 0.6
  ) +
  facet_wrap(~Format, ncol = 3, scales = "free_y") +
  scale_x_continuous(
    breaks = c(1900, 1950, 2000, 2020),
    expand = expansion(mult = c(0.01, 0.06))
  ) +
  labs(
    title = "Format expansion after 1900",
    x = "Year",
    y = "Number of entries"
  ) +
  base_theme +
  theme(legend.position = "none")

p1
```

### Language composition by format {#sec-viola-music-panel-b}

Language profile of each main format, shown as shares within each format.

```{r}
#| label: fig-viola-language-by-format
#| fig-cap: "Language composition by format after 1900"

p2 <- viola_b %>%
  count(Format, Language, name = "n") %>%
  group_by(Format) %>%
  mutate(Share = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(Format, Share, fill = Language)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  scale_fill_manual(
    values = rep(quiet_pal, length.out = n_distinct(viola_b$Language))
  ) +
  labs(
    title = "Language composition by format after 1900",
    x = "Format",
    y = "Share of entries"
  ) +
  base_theme +
  theme(legend.position = "bottom")

p2
```

### Top three languages by format {#sec-viola-music-panel-c}

Absolute counts for the three most common languages within each major
music format; faceting makes it easier to compare language structure
across formats.

```{r}
#| label: fig-viola-top3-by-format
#| fig-cap: "Top three languages by format after 1900"

p3 <- viola %>%
  filter(Language_lab %in% top3_langs) %>%
  count(Format, Language_lab, name = "n") %>%
  group_by(Format) %>%
  mutate(Language_lab = fct_reorder(Language_lab, n)) %>%
  ungroup() %>%
  ggplot(aes(Language_lab, n)) +
  geom_col(fill = "grey35") +
  facet_wrap(~Format, ncol = 3, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Top three languages by format after 1900",
    x = "Language",
    y = "Number of entries"
  ) +
  base_theme +
  theme(legend.position = "none")

p3
```

### Combined focused music figure

```{r}
#| label: fig-viola-music-combined
#| fig-cap: "Music metadata after 1900: format-specific language dynamics"
#| fig-width: 14
#| fig-height: 10

music_figure <-
  (p1 / p2 / p3) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Music metadata after 1900: format-specific language dynamics",
    tag_levels = "a"
  ) &
  theme(
    plot.tag   = element_text(size = 16, face = "bold"),
    plot.title = element_text(size = 18, face = "bold"),
    legend.position = "bottom"
  )

music_figure
```

```{r}
#| label: save-viola-music
#| eval: false

ggsave(
  "viola_music_after1900_abc.png",
  music_figure,
  width = 14,
  height = 10,
  units = "in",
  dpi = 300
)
```
